### 2013-06-06 doc/pgas13/flat-combining.tex
Overhaul of FC section.

Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially. Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure. However, the same delegation mechanism can be reused in any data structure protected by a single global lock.

@@

The benefits of flat-combining can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization. Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.

The story for reducing synchronization is more complicated. The problem with having many threads attempt to acquire a lock is that as contention increases, there will be more failed synchronization attempts (failed \emph{compare and swap}, or \emph{CAS} operations), which must be retried until they succeed.
Instead of having most of the threads repeated fail on acquiring the lock, when threads observe that the lock has already been taken, they can delegate their work to the core with the lock by adding their requests to the structure's publication list. However, in a naive implementation, this operation still needs to perform synchronization every time a request is added to ensure inserts do not collide.
There is no way to avoid all synchronization, however, the cost may be amortized over many operations on the same data structure. Each thread must pay the synchronization cost to insert a publication record into the list once, but after that, it may reuse the record for subsequent accesses without synchronizing.

The final piece of the flat combining story is how operations are semantically combined using knowledge of how data structure accesses compose. All of the threads that did not acquire the lock block on their request until it is satisfied. Whichever thread did obtain the lock becomes the combiner. It iterates over the publication list, merging each non-empty publication record into a combined operation using a combining function defined by the data structure. When the pass is complete, the combined operation is applied to the actual data structure. The combiner thread then takes the aggregated results of the combined operation and distributes them to the correct requests in the publication list. When each request is fulfilled, the thread blocking on it wakes and continues, and finally the combiner thread releases the lock. If any publication requests remain, which could have come from threads that published their request after the combining thread had passed, one of them acquires the lock and becomes the new combiner.

\TODO{Add example of doing flat-combining on a global stack, discuss how pushes/pops are flattened.}

@@

The cost of performing synchronization remotely on a distributed data structure motivates applying the flat-combining paradigm to Grappa data structures. Instead of all of the workers on a core sending messages to do remote synchronization independently, flat-combining can allow all of the workers to share a single remote synchronization operation. In a way, the Grappa runtime is already reducing synchronization traffic when it aggregates small messages that are all going to the same destination. However, it doesn't know what the operations are doing, so it must still serialize, pack, transmit, perform the operation on the remote side, and send a response for each message. Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.

In a PGAS setting, and in Grappa in particular, a number of different choices are made when implementing flat combining. Because memory is physically distributed, locality must be made explicit. In addition, in Grappa there are orders of magnitude more concurrent threads (``workers'') accessing shared data structures. Typically around a thousand are needed per core to tolerate the latency of remote operations, so in a cluster of machines, there are easily millions of workers. Therefore, a different scheme for managing many threads is necessary.

Flat-combined global data structures have a local publication list on each core which serve as proxies to the global structure and are used to combine requests. Workers add their requests to the local publication list by issuing operations on their local proxy and then block waiting for their requests to be satisfied. Similar to the shared-memory case, one requesting worker is chosen to combine requests and perform the combined operation globally. This worker walks the local publication list, combining requests into a local request object as it goes. If it is possible to service a request locally, as it is on some data structures such as matching pushes and pops to a stack, the combiner can satisfy the requests and wake the correct workers even before synchronizing globally. The combiner then issues the combined data structure operation globally. In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency. In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops. On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.

At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs). This would likely result in a net benefit due to local synchronization being inherently cheaper. However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.

By design, each core in Grappa operates independently of the other cores. Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call. Therefore workers within a core can cooperate without any explicit synchronization operations requiring memory fences. This allows them to publish requests to the local  proxy very inexpensively, eliminating the need for a complex synchronization-amortizing queue. This has the added benefit that there will be no unused publication records, which could have become a problem with thousands of workers. Further, remote operations are processed through the same cooperative-multithreading scheduler on the remote side, so combined operations are trivially serialized and atomic.

@@

\section{FC Framework}
In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.

>>>>

At the most basic level, the concept of flat combining is about enabling cooperation among threads rather than contention. The benefits can be broken down into three components: improved locality, reduced synchronization, and data structure-specific optimization. We will explore how this works in a traditional shared memory system, and then describe how the same concepts can be applied to distributed memory.

% Flat combining allows threads to cooperate when accessing a synchronized shared data structure to get greater overall throughput than even each of them operating serially. However, the same delegation mechanism can be reused in any data structure protected by a single global lock.

@@

One of the issues with shared data structures in a shared-memory multicore is that threads executing on different cores will force the hottest parts of the data structure to thrash between their caches. Having a single thread do all of the operations on the data structure allows it to keep everything in its cache.

% Improving locality in a shared-memory multicore is achieved by 
% Having a single thread bound to a particular core do a string of accesses trivially results in a lower cache miss rate than having multiple threads on multiple cores bring the data structure into cache each in turn.

% Careful engineering must be employed to implement a mechanism for cooperation that does not introduce the same synchronization issues as the original contended data structure.

Reduced synchronization comes from delegating work to another core.
To illustrate the issue, consider the shared synchronous queue, shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock. Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock. When many threads contend for the lock, all but one will fail and will have to retry. Each try forces a memory fence for consistency and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets. Under flat combining, instead, threads add requests to the publication list. They each try to acquire the lock, and the one that succeeds becomes the combiner. Instead of retrying, the rest spin waiting for their request to be filled. The combiner walks the publication list, performs all of the requests, and when done, releases the lock. This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization---threads must synchronize to add to the publication list. However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.

The above example of delegation is compelling in itself. However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently than separately.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills each of their records and allows each of them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the remaining two elements are pushed onto the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be combined for greater performance.

@@

In a PGAS setting, and in Grappa in particular, the cost of global synchronization and the amount of concurrency is even greater than in shared memory. With thousands of workers per core, in a reasonably sized cluster there are easily millions of workers. This presents an opportunity for flat combining to pay off greatly, but also poses new challenges.

% a number of different choices are made when implementing flat combining. Because memory is physically distributed, locality must be made explicit. In addition, in Grappa there are orders of magnitude more concurrent threads (``workers'') accessing shared data structures. Typically around a thousand are needed per core to tolerate the latency of remote operations, so in a cluster of machines, there are easily millions of workers. Therefore, a different scheme for managing many threads is necessary.

To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure might look like. Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier. A storage array is allocated out of the global heap, so its elements are striped across all the cores in the system. One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read()} delegate described earlier, blocks the task until complete. To do a \texttt{push}, the task must send a message to the master to acquire the lock. If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All other workers performing operations block at the first message until the lock is released. Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait. However, all thousand workers on each core all perform this synchronization and serialize on the master core.

With flat combining, instead of all workers on a core sending independent messages to do remote synchronization, they can do combining locally and share a single remote synchronization operation. 
As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and as bonus, allows for data structure-specific combining to magnify the benefits even further.

% allows synchronization to happen in parallel on all the cores, 
% Flat combining in Grappa is about allowing synchronization and  distributing synchronization 
% In a way, Grappa's automatic message aggregation is already trying to reduce synchronization traffic; however, the generic mechanism for serializing and packing has a cost, and in the end, the operations must still serialize on the master.
% Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.

% In order for workers to cooperate locally, a local ``proxy'' object is allocated in the core-private heap. This proxy serves as the

When a flat-combined global data structure is created, each core allocates a local proxy object in it's core-local heap. This proxy serves the role of the publication list, allowing workers within a core to coordinate their efforts. 
Conceptually, workers add their requests to the local publication list, and one is chosen to do the combining. However, because workers are scheduled cooperatively, each worker has atomic access to the local proxy ``for free'', so an explicit publication list is unnecessary.
Instead, each worker performs its own combining, and blocks if it was unable to satisfy its request immediately. Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops can be matched locally, avoiding global synchronization for those operations.

After all local combining has been done, one requesting worker is chosen to perform the combined operation globally. In the same figure, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization. As in the uncombined case, the global lock must still be acquired, and concurrent combined requests must still block and serialize on the master, though a technique later will describe how flat combining can be applied here, too. Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.

% In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency. In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops. On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.

% At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs). This would likely result in a net benefit due to local synchronization being inherently cheaper. However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.

% By design, each core in Grappa operates independently of the other cores. Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call. Therefore workers within a core can cooperate without any explicit synchronization operations requiring memory fences. This allows them to publish requests to the local proxy very inexpensively, eliminating the need for a complex synchronization-amortizing queue. This has the added benefit that there will be no unused publication records, which could have become a problem with thousands of workers. Further, remote operations are processed through the same cooperative-multithreading scheduler on the remote side, so combined operations are trivially serialized and atomic.

@@

\section{FC Framework}
Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.

In order to leverage the flat-combining paradigm in Grappa, we implemented a generic flat-combining framework which can be used to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.
